---
title: "Replication code for 'Effects of Starlink adoption on forest degradation in the Amazon' (2026)"
subtitle: "by Borges, Komatsu, Maturano, Nishida, and Menezes Filho"
output: 
  html_document: 
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: false

date: "`r format(Sys.Date(), '%Y-%m-%d')`"
---

<style type="text/css">
.main-container {
  max-width: 100% !important;
  margin: auto;
}
</style>


```{r Initialization, include=FALSE}
# Initializing code
rm(list=ls()); gc()
options(dplyr.summarise.inform = FALSE)
try(Sys.setlocale("LC_CTYPE", "pt_BR.UTF-8"), silent = TRUE) # for special chars

# Loading required R packages 
library(fixest)
library(tidyverse)
library(ggpubr)

# Setting the working directory (root)
setwd(".../code")

# DETER filename (for creating figure)
DETER_zip = "deter-amz-public-2025set01.zip"

# Function for turning first character to upper case
first_upper = function(string) {
  substr(string, 1, 1) = toupper(substr(string, 1, 1))
  return(string)
  }
```


---

# Fixed-Effects IV Regression

```{r DataNormalization, echo=F}
# Dataset and normalization
dta = haven::read_dta("../data/processed/analytical_dataset.dta") %>% 
  filter(year >= 2017 & year <= 2024 & !is.na(coverarea)) %>%
  mutate(year = factor(year, levels = as.character(2024:2017))) %>%
  fastDummies::dummy_cols("year") %>%
  mutate(ca_2017 = year_2017 * coverarea,
         ca_2018 = year_2018 * coverarea,
         ca_2019 = year_2019 * coverarea,
         ca_2020 = year_2020 * coverarea,
         ca_2021 = year_2021 * coverarea,
         ca_2022 = year_2022 * coverarea,
         ca_2023 = year_2023 * coverarea,
         ca_2024 = year_2024 * coverarea,
         sh_nonflo = 1 - sh_flo,
         ) %>%
  select(-starts_with("year_20"))


# Creating new variables / Normalizing by area
area_list = colnames(dta)[startsWith(colnames(dta),"area")]
n_list = colnames(dta)[startsWith(colnames(dta),"n_")]
sh_list = colnames(dta)[startsWith(colnames(dta),"sh_")]

for (y in area_list) {
  dta[[paste0("r_", y)]] = (dta[[y]] / dta[["areamun"]]) * 100
}

for (y in n_list) {
  dta[[paste0("r_", y)]] = (dta[[y]] / dta[["areamun"]]) * 1000
  }

for (y in sh_list) {
  dta[[y]] = dta[[y]] * 100
}

dta = dta %>% mutate(
  areamun = areamun / 1000,
  areaforest = areaforest / 1000,
  dist_coast = dist_coast / 1000,
  dist_brasilia = dist_brasilia / 1000,
) %>% select(-r_areamun)
```


```{r VariableNames, echo=F}
# Defining variables and names
var_names = rbind(
  # Outcomes
  c("r_area_mun_deter_degr","Degradation (% munic. area)", "out"), 
      c("r_area_mun_deter_degr_fire","   fire scars (% munic. area)", "out"),
      c("r_area_mun_deter_degr_wood","   selective-cutting (% munic. area)", "out"),
      c("r_area_mun_deter_degr_outr","   other means (% munic. area)", "out"),
  
  c("r_n_mun_deter_degr","Degradation (flags per 1,000 km²)", "out"),
      c("r_n_mun_deter_degr_fire","   fire scars (flags per 1,000 km²)", "out"),
      c("r_n_mun_deter_degr_wood","   selective-cutting (flags per 1,000 km²)", "out"),
      c("r_n_mun_deter_degr_outr","   other means (flags per 1,000 km²)", "out"),
  
  c("sh_nonflo","Forest replaced (km²)", "out"),
      c("sh_veg", "   by natural non-forest area", "out"),
      c("sh_agr_agr", "   by cropland", "out"),
      c("sh_agr_pec", "   by pasture", "out"),
      c("sh_agr_out", "   by mosaic/forest plantation", "out"),
      c("sh_nvg_agu", "   by water", "out"),
      c("sh_nvg_urb", "   by urban infrastructure", "out"),
      c("sh_nvg_min", "   by mining", "out"),
      c("sh_nvg_out", "   by other non-vegetated area", "out"),
  
  c("r_n_mun_police","Enforcement fines (per 1,000 km²)", "out"),
      c("r_n_mun_police_faun","   for fauna damage", "out"),
      c("r_n_mun_police_polu","   for pollution", "out"),
      c("r_n_mun_police_admi","   for admin. issues", "out"),
      c("r_n_mun_police_flor","   for flora damage", "out"),
      c("r_n_mun_police_defo","   degradation/deforestation", "out"),
      c("r_n_mun_police_fire","   use of fire", "out"),
  
  c("pm10mun","Coarse particulate matter (PM10)", "out"),
  c("pm25mun","Fine particulate matter (PM2.5)", "out"),
  c("pm1mun","Submicron particulate matter (PM1)", "out"),
  
  c("mort_total","Deaths per 100,000 inhab.", "out"),
      c("mort_extcause","   external cause", "out"),
          c("hom","      homicide", "out"),
          c("hom_fa","      homicide by fire arm", "out"),
      c("mort_infect", "   infection", "out"),
      c("mort_neoplasm", "   neoplasm", "out"),
      c("mort_blood", "   blood disorders", "out"),
      c("mort_endocrine", "   endocrine/metabolic issues", "out"),
      c("mort_mental", "   mental disorders", "out"),
      c("mort_nervous", "   nervous system", "out"),
      c("mort_eyes", "   eye conditions", "out"),
      c("mort_ears", "   ear conditions", "out"),
      c("mort_circulatory", "   circulatory issues", "out"),
      c("mort_respiratory", "   respiratory issues", "out"),
      c("mort_digestive", "   digestive issues", "out"),
      c("mort_skin", "   skin conditions", "out"),
      c("mort_musculoskeletal", "   musculoskeletal issues", "out"),
      c("mort_genitourinary", "   genitourinary issues", "out"),
      c("mort_pregnancy", "   pregnancy/childbirth", "out"),
      c("mort_perinatal", "   perinatal causes", "out"),
      c("mort_malformation", "   malformations", "out"),
      c("mort_other", "   other diseases", "out"),
    
  # Covariates  
  c("coverarea", "Mobile coverage (% munic. area)", "cov"),
  c("starlink", "Starlink per 1,000 inhab.", "cov"),
  c("geosat", "GEO internet per 1,000 inhab.", "cov"),
  c("areamun", "Municipal area (1,000 km²)", "cov"),
  c("areaforest", "   Forested area (1,000 km²)", "cov"),
  c("soyield_ful_mun_high", "Soy potential w/ high-tech (ton/ha)", "cov"),  
  
  c("lat","Latitude", "cov"),
  c("lon","Longitude", "cov"),
  c("pop","Population (1,000)", "cov"),
  c("dens","Population density (pop./km²)", "cov"),
  c("dist_coast","Distance to coast (1,000 km)", "cov"),
  c("dist_brasilia","Distance to Brasilia (1,000 km)", "cov"),
  c("tmax","Maximum temperature (ºC)", "cov"),
  c("prcp","Annual precipitation (mm)", "cov")
  
  ) %>% as.data.frame()

colnames(var_names) = c("var", "varname", "type")

varlist = var_names$var[!grepl("prodes", var_names$var, ignore.case = TRUE) &
                          var_names$type == "out"]

```



```{r DescriptiveStatistics, R.options = list(width = 1200), echo=F}
# Descriptive statistics
vars = c("r_area_mun_deter_degr", "r_area_mun_deter_degr_fire",
         "r_area_mun_deter_degr_wood","r_area_mun_deter_degr_outr",
         
         "r_n_mun_deter_degr", "r_n_mun_deter_degr_fire",
         "r_n_mun_deter_degr_wood","r_n_mun_deter_degr_outr",
         
         "r_n_mun_police", 
         "r_n_mun_police_admi", "r_n_mun_police_polu", "r_n_mun_police_faun", 
         "r_n_mun_police_flor", "r_n_mun_police_defo", "r_n_mun_police_fire", 
         
         "sh_nonflo", "sh_veg", "sh_agr_agr", "sh_agr_pec", "sh_agr_out",
         "sh_nvg_min", "sh_nvg_agu", "sh_nvg_urb", "sh_nvg_out",
         
         "pm10mun", "pm25mun", "pm1mun",
         "mort_total", "mort_extcause", "mort_neoplasm", "mort_endocrine",
         
         "coverarea", "starlink", "geosat",
         "areamun", "areaforest", 
         "lat", "lon", 
         
         "soyield_ful_mun_high",
         
         "pop", "dens", "dist_coast", "dist_brasilia", "tmax", "prcp"
         )


# function to format mean ± sd
fmt = function(x, digits = 2) {
  if (all(is.na(x))) return(NA_character_)
  m = mean(x, na.rm = TRUE)
  s = sd(x, na.rm = TRUE)
  sprintf(paste0("%.", digits, "f ± %.", digits, "f"), m, s)
}


for (period in list(2017:2021, 2022:2024)) {
  # create empty tab with names in first column and NA for groups
  tab = data.frame(
    Name = var_names$varname[match(vars, var_names$var)],
    MeanSD = NA_character_,
    stringsAsFactors = FALSE
  )
  
  dta2 = dta %>% filter(year %in% as.character(period)) %>%
        mutate(pop = pop / 1000, N = 1)
  i=1
  # fill tab by looping over vars
  for (var in vars) {
    colvals = dta2[[var]]
    # compute per-group formatted strings
    tab$MeanSD[i] = fmt(colvals)
    i = i + 1
  }
  
  tab = rbind(colnames(tab), 
              c("N", nrow(dta2)/length(period)),
              tab)
  colnames(tab) = NULL
  

  assign(paste0("tab_stat_", first(period), "_", last(period)),
         format(tab, justify = "left"))
}
# View(format(tab, justify = "left"))
```


```{r Covariates, echo=F}
covlist = list(
  # Model 1 - No covariates
  c("1"), 
  
  # Model 2 - All covariates
  c("year:lat", "year:lon", "year:dens", "year:areaforest", 
    "year:dist_coast", "year:dist_brasilia", "tmax", "prcp", "geosat")
)

covs = unique(unlist(covlist))[unique(unlist(covlist)) != "1"]
```



```{r ResultsTable, echo=F}
# Setting results table
result = data.frame(matrix(ncol = 1 + length(covlist), 
                           nrow = 2 + 2*length(varlist) + 2 + length(covs),
                           data = ""))
result[1, 1 + ceiling(length(covlist)/2)] = "Starlink"
result[2, 1] = "Outcomes"
result[2, 2:ncol(result)] = paste0("  (", as.roman(1:length(covlist)), ")")


# # NEW
result = cbind(result, data.frame(ci95 = "", t = "", p = ""))
result[2, (ncol(result)-2):ncol(result)] = c("95% CI", "t-stat", "p-val")
# # NEW

i = 3 + 2*length(varlist)
result[i, ] = "-------"
result[i+1, 1] = "Covariates"
i = i + 2


for (cov in covs) {
  result[i, 1] = paste0("   ", cov)
  rownames(result)[i] = cov
  i = i + 1
}

j = 2
for (covs in covlist) {
  if (covs[1] != "1") {
    for (cov in covs) {
      result[cov, j] = "   X"
    }
  }
  j = j + 1
}


i = 3
for (vardep in varlist) {
  rownames(result)[i] = paste0(vardep, "_est")
  rownames(result)[i+1] = paste0(vardep, "_se")
  result[i, 1] = var_names$varname[which(vardep == var_names$var)]
  
  i = i + 2
}

```



- Model I: no covariates
- Model II: all covariates

```{r Estimations, message=F, warning=F, R.options = list(width = 1200), echo=F}
print(paste0("Covariates: ", paste(covs, collapse = ", ")))

bd_fig = data.frame() # Store results to create figures
j = 2
for (covs in covlist) {
  for (y in varlist) {
    model = paste0(y, " ~ ",
                   # Covariates
                   paste0(covs, collapse=" + "),
                   " | codmun + interaction(uf, year)", # fixed-effects
                   " | starlink ~ coverarea:year") %>% as.formula()
    
    est = feols(model, data = dta %>% filter(reg_sample == 1), cluster="codmun")
    tab = broom::tidy(est, conf.int = T, conf.level = .95)
    
    result[paste0(y,"_est"), j] = paste0(sprintf("%.3f", tab$estimate[1]),
                                         case_when(
                                           tab$p.value[1] <= .01 ~ "***",
                                           tab$p.value[1] <= .05 ~ "**",
                                           tab$p.value[1] <= .10 ~ "*",
                                           TRUE ~ ""))
    result[paste0(y,"_se"), j] = paste0("(", sprintf("%.3f", tab$std.error[1]), ")")
    
    if (all(covs == covlist[[length(covlist)]])) {
      
      if (y == varlist[1]) {
        # 1st stage results
        tab_1st = round(summary(est, stage = 1)$coeftable, 4)
        tab_1stA = fitstat(est, ~cd + ivwald)
        tab_1stB = paste0("R² = ", sprintf("%.4f", r2(summary(est, stage = 1))["r2"]))
      }
      
      result[paste0(y, "_est"), "ci95"] = 
        paste0("[", sprintf("%.3f", tab$conf.low[1]), ", ", sprintf("%.3f", tab$conf.high[1]), "]")
      result[paste0(y, "_est"), "t"] = sprintf("%.3f", tab$statistic[1])
      result[paste0(y, "_est"), "p"] = sprintf("%.3f", tab$p.value[1])
      
      aux = broom::tidy(est, conf.int = TRUE, conf.level = 0.95) %>%
        rename(conf.low95 = conf.low, conf.high95 = conf.high) %>%
        left_join(
          broom::tidy(est, conf.int = TRUE, conf.level = 0.90) %>%
            select(term, conf.low, conf.high) %>%
            rename(conf.low90 = conf.low, conf.high90 = conf.high),
          by = "term") %>%
        filter(term == "fit_starlink") %>%
        mutate(term = y)
      
      # Saving results for figures
      bd_fig = bind_rows(bd_fig, aux)
    }
  }
  
  j = j + 1
}
rownames(result) = 1:nrow(result)
colnames(result) = NULL


print("# 2nd Stage # # # # # # # # # #")
print(format(result[,-c((ncol(result)-2):(ncol(result)-1))], justify = "left"))
# View(format(result[,-c((ncol(result)-2):(ncol(result)-1))], justify = "left"))
```



```{r FigDegradation, echo=F, fig.height=2.5, fig.width=8, message=F, warning=F}
aux = bd_fig %>% filter(startsWith(term, "r_") & grepl("_degr", term)) %>%
  mutate(type = case_when(grepl("_area_", term) ~ "area",
                          grepl("_n_", term) ~ "n"))

aux$term = var_names$varname[match(aux$term, var_names$var)]

aux = aux %>%
  mutate(
    term = sub("\\s*\\(.*\\)$", "", term),
    term = gsub("^\\s+|\\s+$", "", term), # removing leading spaces
    term = sub("^by ", "", term),
    term = ifelse(term == "Degradation", "total", term),
    term = first_upper(term)
  ) 

aux$term = factor(aux$term, levels = unique(aux$term[nrow(aux):1]))

# Degradation as % of municipal area
fig_degr_mun_area = ggplot(aux %>% filter(type == "area"),
                           aes(x = term, y = estimate)) +
  geom_hline(yintercept = 0, color = "black", size = 0.5, linetype = "dashed") +
  
  geom_linerange(
    aes(ymin = conf.low95, ymax = conf.high95, colour = "95% CI", group = type),
    position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path"
  ) +
  geom_linerange(
    aes(ymin = conf.low90, ymax = conf.high90, colour = "90% CI", group = type),
    position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path"
  ) +
  scale_colour_manual(
    name   = NULL,
    breaks = c("90% CI", "95% CI"),
    values = c("90% CI" = "grey60", "95% CI" = "grey80"),
    guide  = guide_legend(
      order = 2,
      override.aes = list(shape = NA, linetype = 1, size = 3)
    )
  ) +

  ggnewscale::new_scale_color() +

  geom_point(
    color = "red",
    position = position_dodge(width = 0.5), size = 2,
    key_glyph = "point"
  ) +

  labs(x = NULL, y = NULL, title = "Degradation (% munic. area)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 11.5),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    text = element_text(size = 11.5)
  ) + 
  coord_flip() + theme(legend.position = "right")


# Degradation as flags per 1,000 km²
fig_degr_mun_n = ggplot(aux %>% filter(type == "n"),
                           aes(x = term, y = estimate)) +
  geom_hline(yintercept = 0, color = "black", size = 0.5, linetype = "dashed") +
  
  geom_linerange(
    aes(ymin = conf.low95, ymax = conf.high95, colour = "95% CI", group = type),
    position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path"
  ) +
  geom_linerange(
    aes(ymin = conf.low90, ymax = conf.high90, colour = "90% CI", group = type),
    position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path"
  ) +
  scale_colour_manual(
    name   = NULL,
    breaks = c("90% CI", "95% CI"),
    values = c("90% CI" = "grey60", "95% CI" = "grey80"),
    guide  = guide_legend(
      order = 2,
      override.aes = list(shape = NA, linetype = 1, size = 3)
    )
  ) +

  ggnewscale::new_scale_color() +

  geom_point(
    color = "red",
    position = position_dodge(width = 0.5), size = 2,
    key_glyph = "point"
  ) +

  labs(x = NULL, y = NULL, title = "Degradation (% munic. area)") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 11.5),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    text = element_text(size = 11.5)
  ) + 
  coord_flip() + theme(legend.position = "right")


# Joining both figures
fig_degr_mun = ggarrange(fig_degr_mun_area, 
                         fig_degr_mun_n + theme(axis.text.y  = element_blank()),
                         widths = c(1,.7), common.legend = T, legend = "bottom")

```



```{r FigFines, echo=F, fig.height=3.5, fig.width=6}
aux = bd_fig %>% filter(grepl("police", term)) %>%
  mutate(
    type = case_when(
      grepl("_defo", term) ~ "Degrad./Deforest.",
      grepl("_fire", term) ~ "Fire",
      grepl("_flor", term) ~ "Flora",
      grepl("_faun", term) ~ "Fauna",
      grepl("_admi", term) ~ "Administrative",
      grepl("_polu", term) ~ "Pollution",
      TRUE ~ "Total"),
    type = factor(type, levels = c("Fire", "Degrad./Deforest.", "Flora", "Fauna",
                                   "Administrative", "Pollution", "Total")))

fig_fines_mun = ggplot(aux, aes(x = type, y = estimate)) +
  geom_hline(yintercept = 0, color = "black", size = 0.5, linetype = "dashed") +

  geom_linerange(
    aes(ymin = conf.low95, ymax = conf.high95, colour = "95% CI"),
    position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path" 
  ) +
  geom_linerange(
    aes(ymin = conf.low90, ymax = conf.high90, colour = "90% CI"),
    position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path" 
  ) +
  scale_colour_manual(
    name   = NULL,
    breaks = c("90% CI", "95% CI"),
    values = c("90% CI" = "grey60", "95% CI" = "grey80"),
    guide  = guide_legend(
      order = 2,
      override.aes = list(shape = NA, linetype = 1, size = 3)
    )
  ) +

  ggnewscale::new_scale_color() +

  geom_point(
    color = "red",
    position = position_dodge(width = 0.5), size = 2,
    key_glyph = "point"   
  ) +

  labs(x = NULL, y = NULL, title = "Fines per 1,000 km²") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 11.5),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    text = element_text(size = 11.5)
  ) + 
  coord_flip() + theme(legend.position = "bottom")

```




```{r FigPollutionDeath, echo=F, fig.height=4.5, fig.width=8}
aux = bd_fig %>% filter(startsWith(term, "pm") | startsWith(term, "mort_")) %>%
  mutate(type = factor(ifelse(startsWith(term, "pm"), "Pollutant (µg per m³)", "Deaths per 100,000 inhab."),
                       levels = c("Pollutant (µg per m³)", "Deaths per 100,000 inhab.")),
         term = ifelse(grepl("pm25", term), "pm25mun", term))

aux$term = var_names$varname[match(aux$term, var_names$var)]

aux = aux %>%
  mutate(
    term = gsub("^\\s+|\\s+$", "", term), # removing leading spaces
    term = sub("^by ", "", term),
    term = sub(" issues", "", term),
    term = sub(" disorders", "", term),
    term = sub(" conditions", "", term),
    term = sub(" causes", "", term),
    term = ifelse(term == "Deaths per 100,000 inhab.", "Total deaths", term),
    term = first_upper(term),
    term = ifelse(term == "Coarse particulate matter (PM10)", "PM10", term),
    term = ifelse(term == "Fine particulate matter (PM2.5)", "PM2.5", term),
    term = ifelse(term == "Submicron particulate matter (PM1)", "PM1", term),
  ) 

aux$term = factor(aux$term, levels = unique(aux$term)[length(unique(aux$term)):1])


fig_pollut = ggplot(aux, aes(x = term, y = estimate)) +
  geom_hline(yintercept = 0, color = "black", size = 0.5, linetype = "dashed") +

  geom_linerange(
    aes(ymin = conf.low95, ymax = conf.high95, colour = "95% CI"),
    position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path"
  ) +
  geom_linerange(
    aes(ymin = conf.low90, ymax = conf.high90, colour = "90% CI"),
    position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path"
  ) +
  scale_colour_manual(
    name   = NULL,
    breaks = c("90% CI", "95% CI"),
    values = c("90% CI" = "grey60", "95% CI" = "grey80"),
    guide  = guide_legend(
      order = 2,
      override.aes = list(shape = NA, linetype = 1, size = 3)
    )
  ) +

  ggnewscale::new_scale_color() +

  geom_point(
    color = "red",
    position = position_dodge(width = 0.5), size = 2,
    key_glyph = "point"
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    text = element_text(size = 11.5)
  ) +
  coord_flip() +
  facet_wrap(~ type, scales = "free",
             ncol = 2
             )
```


```{r FigForestConversion, echo=F, fig.height=4, fig.width=4.5}
aux = bd_fig %>% filter(startsWith(term, "sh_"))

aux$term = var_names$varname[match(aux$term, var_names$var)]

aux = aux %>%
  mutate(
    term = sub("\\s*\\(.*\\)$", "", term),
    term = gsub("^\\s+|\\s+$", "", term), # removing leading spaces
    term = sub("^by ", "", term),
    term = ifelse(term == "Forest replaced", "total", term),
    term = first_upper(term)
  ) 

aux$term = factor(aux$term, levels = aux$term[length(aux$term):1])


fig_forest_conv = ggplot(aux, aes(x = term, y = estimate)) +
  geom_hline(yintercept = 0, color = "black", size = 0.5, linetype = "dashed") +

  geom_linerange(
    aes(ymin = conf.low95, ymax = conf.high95, colour = "95% CI"),
    position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path"
  ) +
  geom_linerange(
    aes(ymin = conf.low90, ymax = conf.high90, colour = "90% CI"),
    position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path"
  ) +
  scale_colour_manual(
    name   = NULL,
    breaks = c("90% CI", "95% CI"),
    values = c("90% CI" = "grey60", "95% CI" = "grey80"),
    guide  = guide_legend(
      order = 2,
      override.aes = list(shape = NA, linetype = 1, size = 3)
    )
  ) +

  ggnewscale::new_scale_color() +

  geom_point(
    color = "red",
    position = position_dodge(width = 0.5), size = 2,
    key_glyph = "point"   
  ) +

  labs(x = NULL, y = "% forested area in previous year") +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    text = element_text(size = 11.5)
  ) +
  coord_flip() + theme(legend.position = "bottom")
```


```{r PlaceboTest, message=F, warning=F, fig.height=3.5, fig.width=5, echo=F}
# Placebo/Event study tests
varlist = list("r_area_mun_deter_degr",  "r_n_mun_deter_degr")

covs = covlist[[length(covlist)]] # full set of covariates

for (var in varlist) {
  tab = tibble()
  
  model = paste0(var, " ~ ", 
                 "ca_2017 + ca_2018 + ca_2019 + ca_2020 + ca_2021 + ca_2023 + ca_2024 + ",
                 paste(covs, collapse = " + "),
                 " | codmun + interaction(uf, year)") %>%
    as.formula()
  
  est = feols(model, data = dta, cluster = "codmun")
  est = broom::tidy(est, conf.int = TRUE, conf.level = 0.95) %>%
    rename(conf.low95 = conf.low, conf.high95 = conf.high) %>%
    left_join(
      broom::tidy(est, conf.int = TRUE, conf.level = 0.90) %>%
          select(term, conf.low, conf.high) %>%
          rename(conf.low90 = conf.low, conf.high90 = conf.high),
        by = "term") %>% 
    select(term, estimate, starts_with("conf.")) %>%
    mutate(term = as.numeric(gsub("ca_", "", term))) %>%
    filter(!is.na(term)) %>%
    rbind(tibble(term = 2022, estimate = 0, conf.low95 = 0, conf.high95 = 0,
                 conf.low90 = 0, conf.high90 = 0)) %>% 
    arrange(term) %>%
    rename(year = term)
    
  tab = bind_rows(tab, est)
  
  var_name = var_names$varname[which(var == var_names$var)]
  var_name = gsub("^\\s+|\\s+$", "", var_name) # removing leading spaces
  var_name = ifelse(grepl("degr_", var), paste0("Degradation ", var_name), var_name)
  var_name = ifelse(grepl("defo_", var), paste0("Deforestation ", var_name), var_name)
  
  
  assign(
    paste0("es_", var, "_mun"),
    ggplot(tab, aes(x = year, y = estimate)) +
    geom_hline(yintercept = 0, color = "black", size = 0.5, linetype = "dashed") +
    geom_vline(xintercept = 2022, color = "gray90", size = 1) +

    geom_linerange(
      aes(ymin = conf.low95, ymax = conf.high95, colour = "95% CI"),
      position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path"
    ) +
    geom_linerange(
      aes(ymin = conf.low90, ymax = conf.high90, colour = "90% CI"),
      position = position_dodge(width = 0.5), size = 2, show.legend = TRUE, key_glyph = "path"
    ) +
    scale_colour_manual(
      name   = NULL,
      breaks = c("90% CI", "95% CI"),
      values = c("90% CI" = "grey60", "95% CI" = "grey80"),
      guide  = guide_legend(
        order = 2,
        override.aes = list(shape = NA, linetype = 1, size = 3)
      )
    ) +

    ggnewscale::new_scale_color() +
    
    geom_point(position = position_dodge(width = 0.5), size = 1.5, color = "red") +
    labs(x = "Year", y = "Estimate", color = "Group") +
    theme_minimal() +
    theme(text = element_text(size = 11.5), 
          panel.grid.minor.y = element_blank())
  )
  
}

# Joining both placebos
fig_es_degr = ggarrange(
  es_r_area_mun_deter_degr_mun + labs(title = "a", y = "% munic. area"),
  es_r_n_mun_deter_degr_mun + labs(title = "b", y = "flags per 1,000 km²"),
  ncol = 1, nrow = 2,
  common.legend = TRUE, legend = "right")
```


## Heterogeneity
Splits observations in half based on a grouping variable (high x low) and estimate results separately.

```{r FigHeterogeneity, message=F, warning=F, R.options = list(width = 1200), echo=F}
varlist = var_names$var[grepl("degr", var_names$var)]
bd_fig = data.frame() # Store results to create figures

{ # Results table
  result = data.frame(matrix(ncol = 1 + length(covlist), 
                             nrow = 2 + 2*length(varlist) + 2 + length(covs),
                             data = ""))
  result[1, 1] = ""
  result[2, 1] = "Outcomes"
  result[2, 2:ncol(result)] = paste0("(", as.roman(1:length(covlist)), "-high)")
  
  
  i = 3 + 2*length(varlist)
  result[i, ] = "-------"
  result[i+1, 1] = "Covariates"
  i = i + 2
  
  
  for (cov in covs) {
    result[i, 1] = paste0("   ", cov)
    rownames(result)[i] = cov
    i = i + 1
  }
  
  j = 2
  for (covs in covlist) {
    if (covs[1] != "1") {
      for (cov in covs) {
        result[cov, j] = "   X"
      }
    }
    j = j + 1
  }
  
  i = 3
  for (vardep in varlist) {
    rownames(result)[i] = paste0(vardep, "_est")
    rownames(result)[i+1] = paste0(vardep, "_se")
    result[i, 1] = var_names$varname[which(vardep == var_names$var)]
    i = i + 2
  }
  
  aux = result[2:ncol(result)]
  for (j in 1:ncol(aux)) {
    aux[[j]] = gsub("-high", "-low", aux[[j]])
  }
  
  result = cbind(result, aux)
  result[1, ceiling(ncol(result)/2)] = "Starlink"
}

dta2 = dta %>% mutate(high = soyield_ful_mun_high > median(soyield_ful_mun_high) * 1)


j = 2
for (covs in covlist) {
  for (y in varlist) {
    model = paste0(y, " ~ ",
                   # Covariates
                   paste0(covs, collapse=" + "),
                   " | codmun + interaction(uf, year)", # fixed-effects
                   " | starlink ~ coverarea:year") %>% as.formula()
    
    # HIGH
    est = feols(model, data = dta2 %>% 
                  filter(reg_sample == 1 & high == 1), cluster = "codmun")
    tab = broom::tidy(est)
    
    result[paste0(y,"_est"), j] = paste0(sprintf("%.3f", tab$estimate[1]),
                                         case_when(
                                           tab$p.value[1] <= .01 ~ "***",
                                           tab$p.value[1] <= .05 ~ "**",
                                           tab$p.value[1] <= .10 ~ "*",
                                           TRUE ~ ""))
    result[paste0(y,"_se"), j] = paste0("(", sprintf("%.3f", tab$std.error[1]), ")")
    
    if (all(covs == covlist[[length(covlist)]])) {
      aux = broom::tidy(est, conf.int = TRUE, conf.level = 0.95) %>%
        rename(conf.low95 = conf.low, conf.high95 = conf.high) %>%
        left_join(
          broom::tidy(est, conf.int = TRUE, conf.level = 0.90) %>%
            select(term, conf.low, conf.high) %>%
            rename(conf.low90 = conf.low, conf.high90 = conf.high),
          by = "term") %>%
        filter(term == "fit_starlink") %>%
        mutate(term = y, half = "high")
      bd_fig = bind_rows(bd_fig, aux)
    }
    
    # LOW
    est = feols(model, data = dta2 %>% 
                  filter(reg_sample == 1 & high == 0), cluster = "codmun")
    tab = broom::tidy(est)
    
    result[paste0(y,"_est"), j + length(covlist)] = 
      paste0(sprintf("%.3f", tab$estimate[1]),
             case_when(tab$p.value[1] <= .01 ~ "***",
                       tab$p.value[1] <= .05 ~ "**",
                       tab$p.value[1] <= .10 ~ "*",
                       TRUE ~ ""))
    result[paste0(y,"_se"), j + length(covlist)] = 
      paste0("(", sprintf("%.3f", tab$std.error[1]), ")")
    
    if (all(covs == covlist[[length(covlist)]])) {
      aux = broom::tidy(est, conf.int = TRUE, conf.level = 0.95) %>%
        rename(conf.low95 = conf.low, conf.high95 = conf.high) %>%
        left_join(
          broom::tidy(est, conf.int = TRUE, conf.level = 0.90) %>%
            select(term, conf.low, conf.high) %>%
            rename(conf.low90 = conf.low, conf.high90 = conf.high),
          by = "term") %>%
        filter(term == "fit_starlink") %>%
        mutate(term = y, half = "low")
      bd_fig = bind_rows(bd_fig, aux)
    }
  }
  
  j = j + 1
}

rownames(result) = 1:nrow(result)
colnames(result) = NULL

print("## Heterogeneity by Soy potential w/ high-tech (ton/ha)")
print(summary(dta$soyield_ful_mun_high))
print(format(result, justify = "left"))
print(" ")


# Figure
aux = bd_fig %>% 
  mutate(
    name = var_names$varname[match(term, var_names$var)],
    half = ifelse(half == "high", "High", "Low"),
    name = sub("\\s*\\(.*\\)$", "", name),
    name = gsub("^\\s+|\\s+$", "", name), # removing leading spaces
    name = sub("^by ", "", name),
    name = ifelse(name == "Degradation", "total", name)
  )

ordem = c("Other means", "Selective-cutting", "Fire scars", "Total")

fig_het_soyield_ful_mun_high = 
  ggplot(
    aux %>% filter(grepl("r_area", term) & grepl("degr", term)) %>%
      mutate(name = first_upper(name), name = factor(name, levels = ordem)),
    aes(x = name, y = estimate)
  ) +
  geom_hline(
    yintercept = 0,
    color = "black",
    size = 0.5,
    linetype = "dashed"
  ) +
  
  geom_linerange(
    aes(
      ymin = conf.low95,
      ymax = conf.high95,
      colour = "95% CI",
      group = half
    ),
    position = position_dodge(width = 0.5),
    size = 2,
    show.legend = TRUE,
    key_glyph = "path"
  ) +
  geom_linerange(
    aes(
      ymin = conf.low90,
      ymax = conf.high90,
      colour = "90% CI",
      group = half
    ),
    position = position_dodge(width = 0.5),
    size = 2,
    show.legend = TRUE,
    key_glyph = "path"
  ) +
  scale_colour_manual(
    name   = NULL,
    breaks = c("90% CI", "95% CI"),
    values = c("90% CI" = "grey60", "95% CI" = "grey80"),
    guide  = guide_legend(
      order = 2,
      override.aes = list(
        shape = NA,
        linetype = 1,
        size = 3
      )
    )
  ) +
  
  ggnewscale::new_scale_color() +
  
  geom_point(
    aes(color = half),
    position = position_dodge(width = 0.5),
    size = 2,
    key_glyph = "point"
  ) +
  scale_color_manual(
    name   = NULL,
    values = c("High" = "red", "Low" = "black"),
    guide  = guide_legend(
      order = 1,
      override.aes = list(
        linetype = 0,
        shape = 16,
        size = 2
      )
    )
  ) +
  
  labs(x = NULL, y = "Degradation (% munic. area)",
       title = "Heterogeneity by Soy potential w/ high-tech (ton/ha)") +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    text = element_text(size = 11.5)
  ) +
  coord_flip() + theme(legend.position = "right") 

```


# Figures and Tables

```{r FigMaps, include=FALSE, message=F, warning=F}
library(sf)

# Starlink subscription rates
starlink = readRDS("../data/processed/starlink.RDS")

# Municipalities polygons
munic_sf = st_read("/vsizip/../data/municipalities_legal_amazon.zip/municipalities_legal_amazon.shp") %>%
  st_make_valid() %>% st_transform(4674) %>% st_make_valid() %>%
  mutate(codmun = as.numeric(geocodigo)) %>%
  select(codmun) %>% merge(c(2022, 2024)) %>% rename(year = y) %>%
  left_join(starlink, by = c("codmun", "year")) %>%
  mutate(starlink = ifelse(is.na(starlink), 0, starlink))


# Mobile Coverage Shapefile
# Aggregate KML files and export as GEOPKG
# List of KML files inside a ZIP (CRS: WGS 84)
  # https://dados.gov.br/dados/conjuntos-dados/cobertura_movel
uf_list = c("ac","am","ap","ma","mt","pa","ro","rr","to")

coverage = data.frame()
for (UF in uf_list) {
  # read KML without extracting it and bind with kml_uf
  coverage = rbind(
    coverage,
    st_read(paste0("/vsizip/../data/areas_cobertas.zip/todas_todas_",
                   UF,"_municipio_simple.kml")) %>%
      mutate(sigla_uf=toupper(UF)))
}

coverage = coverage %>% st_make_valid() %>% 
  st_transform(4674) %>% st_make_valid() %>%
  st_intersection(munic_sf %>% mutate(class = "BR") %>% group_by(class) %>%
                    summarise(geometry = st_union(geometry))) 

# Degradation Shapefile
degrad = st_read(paste0("/vsizip/../data/", DETER_zip, "/deter-amz-deter-public.shp")) %>%
  st_make_valid() %>% rename(class = CLASSNAME, codmun=GEOCODIBGE)  %>%
    mutate(year = year(VIEW_DATE),
           codmun = as.numeric(codmun),
           class = case_when(
             class=="CICATRIZ_DE_QUEIMADA" ~ "deter_degr_fire",
             class=="DEGRADACAO" ~ "deter_degr_outr",
             class=="CS_DESORDENADO" ~ "deter_degr_wood",
             class=="CS_GEOMETRICO" ~ "deter_degr_wood",
             class=="CORTE_SELETIVO" ~ "deter_degr_wood",
           )) %>%
    filter(year %in% c(2022, 2024) & !is.na(class)) %>%
  select(codmun, year, class)

years = c(2022, 2024)
breaks = quantile(munic_sf$starlink[munic_sf$year == 2024],
                  probs = c(0, .2, .4, .6, .8, 1)) # quintiles in 2024

labels = c("[0.0 - 0.5]", "(0.5 – 2.0]", "(2.0 – 5.0]", "(5.0 – 9.8]", "(9.8 – 112.2]")
cols = c(
  "#FFFFFF",  # white for zero
  "#D8D8D8",  # light
  "#AEAEAE",  # light–mid
  "#5F5F5F",  # mid–dark
  "#2B2B2B"   # very dark
)

col_palette = setNames(cols, labels)
col_na = "#E0E0E0"    # color for NA / no data

  
# Creating dummy polygon to show missing values in legend
make_poly = function(shp, x = -57, y = 5, year = 2022, value = 0) {
  crs_map = st_crs(shp)
  
  # point in map CRS
  pt = st_as_sf(data.frame(x = x, y = y), coords = c("x", "y"), crs = crs_map)
  
  # small buffer in meters -> MULTIPOLYGON -> back to map CRS
  mp = pt %>%
    st_transform(3857) %>%
    st_buffer(10) %>%
    st_make_valid() %>%
    st_cast("MULTIPOLYGON") %>%
    st_transform(crs_map)
  
  # attach attributes (ensure factor has all levels)
  mp = mp %>% mutate(codmun = 9999999, year = year, starlink = value)
  mp
}

munic_sf = bind_rows(munic_sf, 
                     make_poly(munic_sf, year=2022, value=56), 
                     make_poly(munic_sf, year=2022, value=7), 
                     make_poly(munic_sf, year=2022, value=4), 
                     make_poly(munic_sf, year=2022, value=1)) %>%
  mutate(starlink_bin = factor(
    cut(starlink, breaks = breaks, labels = labels,
        include.lowest = TRUE, right = TRUE),
    levels = labels)
    )


name = "degr"
loss1 = degrad %>% filter(year == 2022)
loss2 = degrad %>% filter(year == 2024)

starlink_degr_2022 = ggplot(munic_sf %>% filter(year == 2022)) +
  geom_sf(
    aes(fill = starlink_bin),
    color = "grey20",
    size = 0.05,
    alpha = .8
  ) +
  scale_fill_manual(
    name = "Starlink per 1,000",
    values = col_palette,
    na.value = col_na,
    limits = labels,
    drop = FALSE
  ) +
  
  ggnewscale::new_scale_fill() +
  
  # 2) Coverage:
  geom_sf(data = coverage,
          aes(fill = "Mobile"),
          alpha = 0.9,
          color = NA) +
  scale_fill_manual(
    name   = NULL,
    values = c("Mobile" = "blue"),
    breaks = "Mobile",
    labels = "Mobile",
    guide  = guide_legend(override.aes = list(alpha = 0.9, color = NA))
  ) +
  
  ggnewscale::new_scale_fill() +
  
  # 3) Degradation
  geom_sf(
    data = loss1,
    aes(fill = name),
    alpha = .5,
    color = "red",
    linewidth = 0.1
  ) +
  scale_fill_manual(
    name   = NULL,
    values = setNames("red", name),
    breaks = name,
    labels = name,
    guide  = guide_legend(override.aes = list(alpha = .9, color = NA))
  ) +
  
  theme_minimal() + coord_sf(datum = NA) +
  theme(
    text = element_text(size = 11),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8.5),
    legend.key.height = unit(4, "mm"),
    legend.key.width = unit(5, "mm"),
    legend.box.spacing = unit(1, "mm"),
    legend.margin = margin(0, 0, 0, 0),
    plot.margin = margin(0, 0, 0, 0)
  ) +
  annotate(
    "text",
    x = -57,
    y = 5,
    label = 2022,
    size = 3.5,
    hjust = 0.5
  )
  

starlink_degr_2024 = ggplot(munic_sf %>% filter(year == 2024)) +
  geom_sf(
    aes(fill = starlink_bin),
    color = "grey20",
    size = 0.05,
    alpha = .8
  ) +
  scale_fill_manual(
    name = "Starlink per 1,000",
    values = col_palette,
    na.value = col_na,
    limits = labels,
    drop = FALSE
  ) +
  
  ggnewscale::new_scale_fill() +
  
  # 2) Coverage:
  geom_sf(data = coverage,
          aes(fill = "Mobile"),
          alpha = 0.9,
          color = NA) +
  scale_fill_manual(
    name   = NULL,
    values = c("Mobile" = "blue"),
    breaks = "Mobile",
    labels = "Mobile",
    guide  = guide_legend(override.aes = list(alpha = 0.9, color = NA))
  ) +
  
  ggnewscale::new_scale_fill() +
  
  # 3) Degradation
  geom_sf(
    data = loss2,
    aes(fill = name),
    alpha = .5,
    color = "red",
    linewidth = 0.05
  ) +
  scale_fill_manual(
    name   = NULL,
    values = setNames("red", name),
    breaks = name,
    labels = name,
    guide  = guide_legend(override.aes = list(alpha = .9, color = NA))
  ) +
  
  theme_minimal() + coord_sf(datum = NA) +
  theme(
    text = element_text(size = 12),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8.5),
    legend.key.height = unit(4, "mm"),
    legend.key.width = unit(5, "mm"),
    legend.box.spacing = unit(1, "mm"),
    legend.margin = margin(0, 0, 0, 0),
    plot.margin = margin(0, 0, 0, 0)
  ) +
  annotate(
    "text",
    x = -57,
    y = 5,
    label = 2024,
    size = 3.5,
    hjust = 0.5
  )

fig_degr_map = ggarrange(starlink_degr_2022 + theme(legend.position = "none"), 
                         starlink_degr_2024,
                         ncol = 2, nrow = 1,
                         common.legend = T, legend = "bottom")
```


## Fig. 1

**Internet access and forest degradation in the Brazilian Legal Amazon**

```{r Fig1, echo=F, fig.height=3.4, fig.width=8, dpi=200}
print(fig_degr_map)
```



## Fig. 2

**Placebo tests for forest degradation and mobile coverage**

```{r Fig2, echo=F, fig.height=3.5, fig.width=6, dpi=200}
print(fig_es_degr)
```



## Fig. 3

**Impact of Starlink on Amazonian forest degradation**

```{r Fig3, echo=F, fig.height=2.5, fig.width=8, dpi=200}
p = ggarrange(
  fig_degr_mun_area + labs(y = "Degradation (% munic. area)", title=NULL) +
    theme(plot.margin = margin(l = 5, r = 30)),
  fig_degr_mun_n + labs(y = "Degradation (flags per 1,000 km²)", title=NULL) +
    theme(axis.text.y = element_blank(),
          panel.grid.minor.x = element_line(color = "grey85", linewidth = 0.3)), 
  ncol = 2,
  common.legend = TRUE,
  legend = "right",
  widths = c(1.0, .85)
)

print(
  annotate_figure(p, top = text_grob(paste0("a", paste0(rep(" ", 40), collapse=" "),
                                            "b", paste0(rep(" ", 28), collapse=" ")),
                                     face = "bold", size = 11))
  )

```




## Fig. 4

**Impact of Starlink on environmental enforcement**

```{r Fig4, echo=F, fig.height=3, fig.width=6, dpi=200}
print(
  fig_fines_mun + labs(y = "Fines by infraction type (per 1,000 km²)") +
    labs(title = NULL) + theme(legend.position = "right")
  )
```



## Fig. 5

**Impact of Starlink on forest conversion**

```{r Fig5, echo=F, fig.height=3, fig.width=6, dpi=200}
print(
  fig_forest_conv + labs(y = "Forest cover change (%)") +
    theme(legend.position = "right")
  )
```



## Fig. 6

**Impact on pollution and human health**

```{r Fig6, echo=F, fig.height=4.0, fig.width=8, dpi=200}
print(fig_pollut + labs(title = paste0("     a", paste0(rep(" ", 35), collapse = " "), "b")))
```





## Table S1

**Descriptive statistics (2017-2024).**

```{r TableS1, R.options = list(width = 1200), echo=F}
tab_stat = cbind(
  tab_stat_2017_2021[,1:2],
  tab_stat_2022_2024[,2]
  )
colnames(tab_stat) = NULL
tab_stat[1,] = c("Variables                             ",
                 "2017-2021       ", "2022-2024       ")
tab_stat = tab_stat[!grepl("TI|UC", tab_stat[[1]]),]
print(tab_stat)
```


## Table S2

**First-stage regression results**

```{r TableS2, R.options = list(width = 1200), echo=F}
print(tab_1st)
print(tab_1stA)
print(tab_1stB)
```


## Fig. S1

**Impact of Starlink on forest degradation by potential soybean yield with high-technology**

```{r FigS1, echo=F, fig.height=3, fig.width=6, dpi=200}
print(fig_het_soyield_ful_mun_high + labs(title = NULL))
```




